What should I notice? Evaluating the memorability of events for abductive inference.

### 2021-10-25 new_paper.tex
Ada's comments, pt. 1

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ When confronted to an unprecedented situation, humans typically show good
performance in quickly identifying noticeable past events and proposing them
as possible causal hypotheses. This kind of abductive inference is widely
overlooked in modern AI approaches which rely on massive datasets to learn
associative patterns. Our proposal is to formalize and compute a ``memorability''
score over a set of events recorded from a cyber-physical system.
This score can then be used to select relevant information to be
remembered and to propose causal hypotheses in unusual situations on demand.
The approach is meant to be complementary to more traditional
learning-focused abduction techniques. We provide theoretical ground for our approach
based on Algorithmic Information
Theory. We also provide an implementation example, a
smart-home scenario, to show how our approach could translate in practice.}

@@

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds go down. Intrigued by this behavior, she
quickly infers that both light dimming and blind closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}. This mental operation
is a key element of humans' ability to understand the world: from the
observed consequences, they infer the possible causes.

In this example, there are mainly three possibilities to come to the
conclusion. (1) If the user knows how the smart living-room system works, if she
knows the underlying rules or parameters, she may use this causal
knowledge to perform abduction. (2) If she has no knowledge
about the system but made several observation of the same behavior, she may examine past
correlations and figure out that turning on the TV set often leads the blinds to
close and the lights to dim. (3) If there are no previous occurrences of the event (e.g. it is the first time she turns on the TV
in the living-room), she may still be able to
suspect that the TV is a possible cause for the observed event, just because it appears to her as a memorable recent event (as it is its first occurrence). This example
suggests that human beings are able to use distinct methods to perform abductive tasks
and infer new knowledge. While the first two mechanisms can be
automated using knowledge bases and statistical methods, ``memorability-based'' abduction needs to be investigated in more depth.

% For instance, it is
% possible, for humans without any prior knowledge about
% physics or electrical engineering, to infer a memorable thunderstorm as a
% possible cause for a general black-out just by using the memorability of
% the former. While this kind of abductive inference (i.e. inference to the cause
% of an observation \cite{magnani_abduction_2011}) can yield many
% false-positives, it can be considered in situations where the lack of knowledge
% from previous occurrences penalizes other approaches.

% The difficulty of this approach to reasoning is, however, to quickly identify
% the relevant candidate hypotheses. In this regards, we may use a basic
% instinct: without prior knowledge, the most relevant hypothesis might simply be
% the most memorable recent event. However this definition is highly subjective,
% and does not seem to fit well into the canvas of computing and AI.

Defining a score of memorability is not straightforward. First, events can be of different nature, and not directly
comparable. Even for restricted systems such as smart homes, noticeable events
range from device removal to presence detection or unusually high
temperatures. Even for comparable events, the problem is to weigh different characteristics: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To
our knowledge, no current system proposes to combine various
event types from different devices to compute a unified metrics of
``memorability''.

To address this issue, we started from the following intuition: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of quantitative or qualitative qualifiers, the most memorable ones are
likely to require less words to be described. ``Last year's hottest day'' is simpler than ``the 182\textsuperscript{nd} day 7 years ago''. How can we quantify this relative simplicity? We could evaluate the complexity of each
description, taking into account both the complexity of the concept words (a date
of occurrence, a temperature ranking), and of the arguments (1st hottest,
182 and 7). This approach, The resulting values define the  \emph{description complexity}
of events. Our intuition is that memorable events require simpler and less numerous qualifiers to be
unambiguously described than unremarkable ones.

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{figures/filters}
    \caption{Retrieving an event through successive predicative
        filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (gray arrows). For example,
        filter $f_1$ selects events from last year, i.e. which satisfy the
        predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
        applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
        year's hottest day. The complexity of this event can then be upper-bounded
        by the complexity of the three filters, since they give an unambiguous way
        to describe the event within the memory.}
    \label{fig:filters}
\end{figure}

For machines to implement and compute description complexity, we need a formal framework and computation methods that are in line with human intuition.  Algorithmic Information Theory (AIT) appears to be such a framework, as it is consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being elements stored in what we call \emph{base memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to events. For instance, the predicate $\mathtt{date}
    (\cdot, 1\_year)$ is \emph{true} of events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation. It generates another memory that is a subset of the previous one.
The filtering operation can then be repeated, selecting
fewer events at each iteration, until a singleton memory is reached. This means that the sequence
of predicates could unambiguously \emph{retrieve} the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this process for the event: ``last year's hottest day''.

We will first briefly introduce
some basic notions of Algorithmic Information Theory. Then, in
section~\ref{sec:computing}, we show how we compute the
memorability of past events. We illustrate our approach in~
(section \ref{sec:example}) by describing an implementation
in a smart home simulation context. In section~\ref{sec:related} we review
a few studies that use complexity theory to generate explanations for the smart home
and we discuss how our work can be linked with them. We eventually explore potential
extensions of our work in section \ref{sec:future}

@@

\subsection{Theoretical Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{Though the definition
    holds for some infinite binary strings (think of the representation of the
    decimals of $\pi$), we restrict ourselves here to finite strings.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. The complexity $K(s)$ of a (finite) binary string is the length in bits of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.

\begin{equation}
    \label{eq:kolmo_def}
    K_{U}(s) = \min_{p}\left\{l(p)|U(p)=s\right\}
\end{equation}

@@

\begin{equation}
    \label{eq:unexpected} U(e) = K_{w}(e) - K(e)
\end{equation}

@@

Using these insights from AIT, we define the memorability $M(e)$ of an event as the absolute difference between the description complexity $K_d(e)$ of an event and its expected description complexity $K_{exp}(e)$:

@@

Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: we do so to acknowledge the fact that events more complex than expected can be memorable as well\footnote{In the original paper \cite{dessalles2011coincidences}, exceptionally complex events are described by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we define computational approximations for the description complexity $K_d$ and the expected complexity $K_{exp}$ of events.

@@

\begin{equation}
    \label{eq:event}
    e = (l, t,\mathcal{D})
\end{equation}

where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a multidimensional data point representing the various characteristics of $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

To model how humans are able to describe events by using qualifiers, we use \emph{predicates}: Boolean functions operating on events and, possibly, additional parameters: $\pi(e, a_1, a_2, \dots, a_n) \mapsto \{O,1\}$ is a predicate of arity $n$ operating on event $e$. In the rest of this paper, we will prefer the equivalent notation $\pi(e, k) \mapsto \{0,1\}$, where $k$ is a binary string encoding the sequence of arguments $a_1, \dots, a_n$. Using this notation, the predicate $\pi$ becomes a boolean function operating on $\mathbf{E} \times \{0,1\}^*$:

\begin{equation}
    \label{eq:predicate}
    \pi : \begin{cases}
        \mathcal{M}\times \{0,1\}^{*} & \mapsto \{0,1\}    \\
        (e, k)                        & \mapsto \pi_{k}(e)
    \end{cases}
\end{equation}

@@

As events occur, they are stored in the  \emph{base memory} $M_0$. As they are not directly comparable, the memory $M_0$ can be considered as having the structure of an unordered set. We denote by $\mathcal{M}$ the subsets of $M_0$. By extension, elements of $\mathcal{M}$, i.e. subsets of $M_0$, are also called \emph{memories}.

@@

\begin{equation}
    \label{eq:filter}
    f_{\pi, k}: \begin{cases}
        \mathcal{M} & \mapsto \mathcal{M}           \\
        M           & \mapsto \{e \in M | \pi(e) \}
    \end{cases}
\end{equation}

@@

\begin{align}
    \label{eq:bit_lenght_p}
    L(p) & = L((f_{\pi_1,k_1}, \dots, f_{\pi_n, k_n}))     \\
         & = L(\pi_1) + L(k_1) + \dots + L(\pi_n) + L(k_n)
\end{align}

@@

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting from the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
    {max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
            {explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

@@

$K_{exp}(e)$ evaluates the complexity the user, or the system, would expect the occurrence of event $e$ to have, based on their previous knowledge. In our framework, this prior knowledge consists of the base memory $M_0$. The expected complexity of the event $e$ can be computed with a simple first-order approximation, i.e. estimating the average complexity of ``similar events'' over the base memory $M_0$.

Still, there is a difficulty in defining what
should be considered \emph{similar} events. Given that we deal with non comparable events, we may define the notion of similarity by referring once again to \emph{predicates}.For a given event $e$ and a given predicate $\pi_k$, we define a $\pi_k$-neighborhood of $e$ as the set $N_{\pi, k}(e)$ of all other events satisfying $\pi_k$.

\begin{equation}
    \label{eq:similar}
    N_{\pi, k}(e) = \{e'\in M_0, \pi_k(e') \wedge e' \neq e\}
\end{equation}

Now, when considering, for all possible predicates $\pi_k$, the corresponding
neighborhoods $N_{\pi, k}(e)$, with the convention that $N_{\pi, k}(e) = \emptyset$
if $e$ does not satisfy $\pi_k$, we can compute an average expected complexity for $e$:

@@

This definition is consistent with the intuitive idea that more similar events should weigh more in the computation. Indeed, if $e'$
is very similar to $e$, it will appear in many neighborhoods, since it
satisfies most of the predicates that $e$ satisfies. Therefore, it will
be present in more terms in Equation \ref{eq:expected}, and will weigh more in
the final result.

\subsubsection{From memorability to abduction}
Abductive inference builds on the computation
of the memorability score. \emph{Knowing} that we want to find a cause $c$
for an observed effect $e$, we try to find the most remarkable event in past
memory that is related to $e$. While our ``memorability'' score identifies remarkable past events, it does not take into account their relatedness to $e$.

The knowledge attached to the occurrence of $e$ can be integrated into the description complexity definition by using conditional complexity $K_d(c | e)$: The information contained in $e$ is considered as given, and therefore ``free'' in terms of complexity. For instance, when looking for a cause for
an anomaly in the living-room, other anomalies occurring in the same
living-room will be simpler, as the location ``living-room'' is already known
from the observation of the current anomaly.

Formally, we now consider only predicates where knowledge of the effect event $e$  is appended to all programs $k$:
$\pi_{k::e}(c)$, where $::$ is the append operation. The set of paths obtained with such predicates is noted $P^\infty_e$. This append operation is free in terms of bit-length in the computation of complexity, since the effect event $e$ is an input of the problem. Therefore, we have $L'(\pi_{k::e}) = L(\pi_k) = L(\pi) + L(k)$. We get a definition for the conditional description complexity:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\subsection{Setup}

@@

\begin{table}
    \centering
    \begin{tabular}{l|c|c}
        Event Id & Description & Relative memorability (bits)\\
        \hline
        2513 & Use of smart TV & 16.76 \\
        2427 & Last use of the old TV & 14.81 \\
        2411 & Second-last use of the old TV & 11.21 
    \end{tabular}
    \caption{Output of the memorability-based abduction module: top 3 events for the relative memorability metrics.}
    \label{tab:abduction_res}
\end{table}

@@

Given that we generated the data used for this experiment, it is possible to
flag all perturbation events from the usual daily events and evaluate how a
detection based on ``memorability'' score would perform in distinguishing these
events. The result is presented as a ROC curve in Figure \ref{fig:roc}. This shows the memorability score's performance as a classifying tool for ``out-of-the norm'' events. In this example, memorability alone correctly identified 18 manually generated events with only 2 false-positives. While the direct application of memorability for classification or event detection is not within the scope of this paper (see Section \ref{sec:related} for complexity-based detection), this first result is on par with the motivation of memorability being in accordance with the intuitive notion.

@@

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ When confronted with an unprecedented situation, humans typically show good
performance in quickly identifying noticeable past events and proposing them
as possible causal hypotheses. This kind of abductive inference is widely
overlooked in modern AI approaches which rely on massive datasets to learn
associative patterns. Our proposal is to formalize and compute a ``memorability''
score over a set of events recorded from a cyber-physical system.
This score can then be used to select relevant information to be
remembered and to propose causal hypotheses in unusual situations on demand.
The approach is meant to be complementary to more traditional
learning-focused abduction techniques. We provide a theoretical framework based on Algorithmic Information Theory to define the notion of memorability of events. Then, we implement our approaches in two examples based on smart-home scenario to validate our approach in toy cases and illustrate how it could be applied.}

@@

As a user has just switch on the TV in her all-equipped living-room, the
lights dim and the window blinds go down. Intrigued by this behavior, she
quickly infers that both light dimming and blind closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}. This mental operation
is a key element of humans' ability to understand the world: from the
observed consequences, they infer the possible causes.
% TODO: answer in the next paragraph?

In this example, there are mainly three possibilities to come to the
conclusion. (1) If the user knows how the smart living-room system works, if she
knows the underlying rules or parameters, she may use this causal
knowledge to perform abduction. (2) If she has no knowledge
about the system but made several observation of the same behavior, she may examine past
correlations and figure out that turning on the TV set often leads the blinds to
close and the lights to dim. (3) If there are no previous occurrences of the event (e.g. it is the first time she turns on the TV
in the living-room), she may still be able to
suspect that the TV is a possible cause for the observed event, just because it appears to her as a memorable recent event (as it is its first occurrence). This example
suggests that human beings are able to use distinct methods to perform abductive tasks
and infer new knowledge, depending on the situation. While the first two mechanisms can be
automated using knowledge bases and statistical methods,the third approach, which can be used without any knowledge of the occurring phenomenon or past occurrences. Instead, it relies on the observation that some events appear more ``memorable'' than others and, as such, can be considered as possible hypotheses if need be.

Defining a score of memorability is not straightforward. First, events can be of different nature, and not directly
comparable. Even for restricted systems such as smart homes, noticeable events
range from device removal to presence detection or unusually high
temperatures. Even for comparable events, the problem is to weigh different characteristics: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To
our knowledge, no current system proposes to combine various
event types from different devices to compute a unified metrics of
``memorability''. 
% ! Addiction of the follwoing sentence, to show what we can use memorability for.
In addition to the aforementionned use for abductive inference, having access to a computation of memorability would allow a system to summarize a situation to its user by presenting only the most memorable events: for instance, a summary of notable events that occurred during the home owner's absence.

% ? What to do with the "intuition" thing? Keep it? Leave it?
To address this issue, we started from the following intuition: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of quantitative or qualitative qualifiers, the most memorable ones are
likely to require less words to be described. ``Last year's hottest day'' is simpler than ``the 182\textsuperscript{nd} day 7 years ago''. How can we quantify this relative simplicity? We could evaluate the complexity of each
description, taking into account both the complexity of the concept words (a date
of occurrence, a temperature ranking), and of the arguments (1st hottest,
182 and 7). The resulting values define the  \emph{description complexity}
of events. Our intuition is that memorable events require simpler and less numerous qualifiers to be
unambiguously described than unremarkable ones.
% ? Response to Ada's remark: added a thing in the 1st paragraph to show what we want ot do with the memorable events.

For machines to implement and compute description complexity, we need a formal framework and computation methods that are in line with human intuition.  Algorithmic Information Theory (AIT) appears to be such a framework, as it is consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being elements stored in what we call \emph{base memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to events. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ is \emph{true} of events that occurred last year. Selecting
%%%%%%%%%%%%%%%%%%%%%% ! MOVED FIGURE HERE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[hb]
    \centering
    \includegraphics[width=.95\linewidth]{figures/filters}
    \caption{Retrieving an event through successive predicative
        filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (gray arrows). For example,
        filter $f_1$ selects events from last year, i.e. which satisfy the
        predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
        applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
        year's hottest day. The complexity of this event can then be upper-bounded
        by the complexity of the three filters as they give an unambiguous way
        to describe the event within the memory.}
    \label{fig:filters}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%! FIGURE END %%%%%%%%%%%%%%%%%%%%%%%%%%%%
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation. It generates another memory that is a subset of the previous one.
The filtering operation can then be repeated, selecting
fewer events at each iteration, until a singleton memory is reached. This means that the sequence
of predicates could unambiguously \emph{retrieve} the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this process for the event: ``last year's hottest day''.

% * Paragraph re-written
We will first briefly introduce
some basic notions of Algorithmic Information Theory in subsection~\ref{sec:theory}. We then expose our contribution by presenting in subsection~\ref{sec:computing} our formal definition of memorability and related notions. Then, we present an implementation example of these definitions with two smart-home examples in section~\ref{sec:results}. The results of these experiments are then presented and discussed. Finally we explore other related works in subsection \ref{sec:related} and explore possible extensions of our work in subsection \ref{sec:future}.

% Then, in
% subsection~\ref{sec:computing}, we show how we compute the
% memorability of past events. We illustrate our approach in section~\ref{sec:example} by describing an implementation
% in a smart home simulation context. In section~\ref{sec:related} we review
% a few studies that use complexity theory to generate explanations for the smart home
% and we discuss how our work can be linked with them. We explore potential
% extensions of our work in section \ref{sec:future}

@@

\subsection{Theoretical Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{Though the definition
    holds for some infinite binary strings (think of the representation of the
    decimals of $\pi$), we restrict ourselves here to finite strings.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. The complexity $K(s)$ of a (finite) binary string is the length in bits $L(p)$ of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.

\begin{equation}
    \label{eq:kolmo_def}
    K_{U}(s) = \min_{p}\left\{L(p)|U(p)=s\right\}
\end{equation}

@@

\begin{equation}
    \label{eq:unexpected} \textit{Unex}(e) = K_{w}(e) - K(e)
\end{equation}

@@

As \cite{dessalles2011coincidences} suggests a link between the unexpectedness and cognitive relevance, we propose to define the memorability of an event in a similar way.  
Hence, we define the memorability $M(e)$ of an event as the absolute difference between its description complexity $K_d(e)$ and its expected description complexity $K_{exp}(e)$:

@@

% ? Peut-être ajouter un paragraphe sur pourquoi ne pas directement utiliser la notion de unexpectedness. 1) pas calculable. 2) pas forcément bien définie dans un cas pratique d'application (comment définit la complexité causale ???)
Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: we do so to acknowledge the fact that events more complex than expected can be memorable as well\footnote{In the original paper \cite{dessalles2011coincidences}, exceptionally complex events are described by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we define computational approximations for the description complexity $K_d$ and the expected complexity $K_{exp}$ of events.

@@

\begin{equation}
    \label{eq:event}
    e = (l, t,\mathcal{D})
\end{equation}
% ! response to Ada's remark: there is a hidden unique ID in the implementation, to distinguish between events; however this is not used to retrieve the event, as it is not a defining characteristics.
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a vector of properties characterizing $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

To model how humans are able to describe events by using qualifiers, we use \emph{predicates}: Boolean functions operating on events and, possibly, additional parameters: $\pi(e, a_1, a_2, \dots, a_n) \mapsto \{O,1\}$ is a predicate of arity $n$ operating on event $e$. In the rest of this paper, we will prefer the equivalent notation $\pi(e, k) \mapsto \{0,1\}$, where $k$ is a binary string encoding the sequence of arguments $a_1, \dots, a_n$. Using this notation, the predicate $\pi$ becomes a boolean function operating on $\mathbf{E} \times \{0,1\}^*$, where $\mathbf{E}$ denotes the set of all events:

\begin{equation}
    \label{eq:predicate}
    \pi : \begin{cases}
        \mathbf{E}\times \{0,1\}^{*} & \mapsto \{0,1\}    \\
        (e, k)                        & \mapsto \pi_{k}(e)
    \end{cases}
\end{equation}

@@

% ! response to Ada's remark: complexity computations don't consider time complexity, only program length. Indeed, using a FILO structure for the memory makes sense implementation-wise, but it does not affect the definitions of complexity and memorability. 
As events occur, they are stored in the  \emph{base memory} $M_0$. As they are not directly comparable, the memory $M_0$ can be considered as having the structure of an unordered set. We denote by $\mathcal{M}$ the set of all subsets of $M_0$. By extension, elements of $\mathcal{M}$, i.e. subsets of $M_0$, are also called \emph{memories}.

@@

\begin{equation}
    \label{eq:filter}
    f_{\pi, k}: \begin{cases}
        \mathcal{M} & \mapsto \mathcal{M}           \\
        M           & \mapsto \{e \in M | \pi_k(e) \}
    \end{cases}
\end{equation}

@@

\begin{align}
    \label{eq:bit_lenght_p}
    L(p) & = L((f_{\pi_1,k_1}, \dots, f_{\pi_n, k_n}))     \\
         & = L(\pi_1) + L(k_1) + \dots + L(\pi_n) + L(k_n)
\end{align}
where $L(\pi_\square)$ and $L(k_\square)$ respectively denote the length, in bits, required to express the predicate's concept and program. This length may vary depending of the encoding choice, see Section \ref{sec:results} for an example.

@@

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting from the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
    {max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
            {explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).
% ! response to Ada's comment: yes, this works well ;-) It is even implemented in the code. However, given that time complexity does not impact the definitions of complexity, I decided to present the simplest program, not the most optimized, for the sake of clarity.

@@

$K_{exp}(e)$ evaluates the complexity that the user, or the system, would expect for the occurrence of event $e$ to have, based on their previous knowledge. In our framework, this prior knowledge consists of the base memory $M_0$. The expected complexity of the event $e$ can be computed with a simple first-order approximation, i.e. estimating the average complexity of ``similar events'' over the base memory $M_0$.

Still, there is a difficulty in defining what
should be considered \emph{similar} events. Given that we deal with non comparable events, we may define the notion of similarity by referring once again to \emph{predicates}. For a given event $e$ and a given predicate $\pi_k$, we define a $\pi_k$-neighborhood of $e$ as the set $N_{\pi, k}(e)$ of all other events satisfying $\pi_k$.

\begin{equation}
    \label{eq:similar}
    N_{\pi, k}(e) = \{e'\in M_0, (e' \neq e) \wedge \pi_k(e')\}
\end{equation}

Now, when considering, for all possible predicates $\pi_k$, the corresponding
neighborhoods $N_{\pi, k}(e)$, with the convention that $N_{\pi, k}(e) = \emptyset$
if $e$ does not satisfy $\pi_k$, we can compute an average expected complexity for $e$, by summing the complexity of events in all neighbourhoods of $e$ and dividing the total by the number of events in the neighborhoods:

@@

This definition is consistent with the intuitive idea that more similar events should weigh more in the computation. Indeed, if $e'$
is very similar to $e$, it will appear in many neighborhoods, since it
satisfies most of the predicates that $e$ satisfies. Therefore, it will
be present in more terms in Equation \ref{eq:expected}, and will weigh more in
the final result.
% ! response to Ada's comment: Yes, it is. If a point belongs to many neighbourhoods, with many points in each, it will be hard to distinguish from the other points. So, its expected complexity is hard. Think, for instance, of the expected complexity for a human being: it will be roughly log2(7000000000) bits, for a randomly picked person. If this person is known to you, it will be simpler to describe her.

\subsubsection{From memorability to abduction}
% TODO : see ada's comment for this part
Abductive inference builds upon the computation
of the memorability score. \emph{Knowing} that we want to find a cause $c$
for an observed effect $e$, we try to find the most remarkable event in past
memory that is related to $e$. While our ``memorability'' score identifies remarkable past events, it does not take into account their relatedness to $e$.

The knowledge attached to the occurrence of $e$ can be integrated into the description complexity definition by using conditional complexity $K_d(c | e)$: The information contained in $e$ is considered as given, and therefore ``free'' in terms of complexity. For instance, when looking for a cause of
an anomaly in the living-room, other anomalies occurring in the same
living-room will be simpler, as the location ``living-room'' is already known
from the observation of the current anomaly.

Formally, we now consider that knowledge of the effect is given. This consists, for instance, of appending a description of effect $e$  is appended to all programs $k$:
$\pi_{k::e}(c)$, where $::$ is the append operation. The set of paths obtained with such predicates is noted $P^\infty_e$. This append operation is free in terms of bit-length in the computation of complexity, since the effect event $e$ is an input of the problem. Therefore, we have $L'(\pi_{k::e}) = L(\pi_k) = L(\pi) + L(k)$. We get a definition for the conditional description complexity:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
\subsection{Setup}

@@

\begin{table}
    \centering
    \begin{tabular}{l|c|c}
        Event Id & Description                   & Relative memorability (bits) \\
        \hline
        2513     & Use of smart TV               & 16.76                        \\
        2427     & Last use of the old TV        & 14.81                        \\
        2411     & Second-last use of the old TV & 11.21
    \end{tabular}
    \caption{Output of the memorability-based abduction module: top 3 events for the relative memorability metrics.}
    \label{tab:abduction_res}
\end{table}

@@

Given that we generated the data used for this experiment, it is possible to
flag all perturbation events from the usual daily events and evaluate how a
detection based on ``memorability'' score would perform in distinguishing these
events. The result is presented as a ROC curve in Figure \ref{fig:roc}. This shows the memorability score's performance as a classifying tool for ``out-of-the norm'' events. In this example, memorability alone correctly identified 18 manually generated events with only 5 false-positives. While the direct application of memorability for classification or event detection is not within the scope of this paper (see Section \ref{sec:related} for complexity-based detection), this first result is on par with the motivation of memorability being in accordance with the intuitive notion.

@@

% ? Maybe a small part on the importance of the introduction of the random pick predicate ?
% The stark difference observed between the two examples in this work shows the importance of being able to select a random event from the set of
